\thispagestyle{plain}
\section{Notes On This Edition}

Only hard copies of the original paper now exist. This new edition has been created by scanning the first edition and processing the results using optical character recognition. The document was spell checked and the original layout of the thesis restored as best as possible on \XeLaTeX \ on a Mac OS X system. Diagrams, tables and citations have all been recreated so they can be most easily read.

No intentional copy changes have been made. The aim has been to present the paper as it was originally delivered.

While my strong support on Ridley's position on climate change (\cite{Ridley:1996}) has not stood the test of time, the \textit{bounded rationality} of even a government in dealing with the problem has, \textit{viz.} the mad-rush to diesel (for fuel efficiency) and reversal (for air quality) shows how difficult it is to game markets. 

If there is one lesson from this paper, and subsequent history, it is, policy is just \textit{difficult.}

\textit{October, 2017}

\section{Original Acknowledgements}

I would like to thank my supervisor Professor John Hibbs for both providing references to material and his useful comments. I would also like to thank the libraries of the Institute of Economic Affairs, the Chartered Institute of Transport and the Office of Passenger Rail Franchising for the information they have provided. This document was typeset using \LaTeXe  \ on a GNU/Linux system. The views expressed here are my personal views and should not be attributed to those who have been mentioned above.

\textit{September 21, 1997}